---
title: "r4bs: Data Manipulation"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First things first - let's make sure you have all the necessary files. 

1. Navigate to the [r4bs github repo](https://github.com/erramirez/r4bs)
2. Dowload or Clone the repo
3. Open the r4bs.Rproj file

## Tidy Data
When working with the tidyverse family of packages and analytical processes it's best to follow tidy data princples. In the book, [_R for Data Science_](http://r4ds.had.co.nz/tidy-data.html), tidy data is described as:

> There are three interrelated rules which make a dataset tidy:
> * Each variable must have its own column.
> * Each observation must have its own row.
> * Each value must have its own cell.

However, not all data is created in such a clean and consistent manner. Good thing there are some functions and methods included in the `tidyr` package (loaded with `tidyverse`) that can help turn messy data into tidy data. 
## Transforming 
The two most commonly used functions are `spread()` and `gather()`. These two functions help clean and "tidy" data that is either:
* *spread* across multiple *columns*
* *scattered* across multiple *rows*

Before we go into the actual functions themselves it's important to understand few terms that we're going to use: _key value pair_. For these functions you will need to identify each _key_ and each _value_ pair to properly spread or gather the data. 

* key
* value


### gather()
When you have data that includes variables that are set as column names and you need to gather those columns into neat and tidy rows (remember, each observation has its own row) you can use the `gather()` function. 

Let's examine a data set that has step data spread across many columns:
```{r gather_example_data}
# load tidyverse to start!

library(tidyverse)
gather_example <- read_csv("~/GitHub/r4bs/ExampleData/FitbitDailyData/p1_p5_Steps.csv")

gather_example
```

You can see that for this data file, we have daily step data for five participants, but each column is the date - sometimes called a "wide" format. We want to turn that into a narrow file, with three variables: id, date, and totalsteps. 

```{r gather}

p1_p5_DailySteps<- gather_example %>% 
  gather(`1/9/12`, `1/10/12`, `1/11/12`, `1/12/12`, `1/13/12`, `1/14/12`, `1/15/12`, 
         key = "date", value = "totalsteps")

p1_p5_DailySteps
```

Now we have nice and neat data set with one observation (one day) per row. 

### spread()
As you may have guessed, spreading is the opposite of gathering. When you have data that has multiple rows when it should have multiple columns, we can use `spread()` to tidy it. 

Let's start with some example data. 
```{r spread_example_data}
# date should be in date format so we use col_date here to format it
spread_example <- read_csv("~/GitHub/r4bs/ExampleData/FitbitDailyData/p1_p5_ActivityIntensity.csv", 
                           col_types = cols(
                             date = col_date("%m/%d/%y")
                           )
                           )
```
You can see here that four different activity intensity variables are scattered across multiple rows. We need to get these variables into columns. 

* The column that contains variable names we use as the `key`: `activityintensity`

* The column that contains values for the variables we use as the `value` : `minutes`

```{r spread_example}

p1_p5_DailyIntensityMinutes <- spread_example %>% 
  spread(key = activityintensity, value = minutes)

p1_p5_DailyIntensityMinutes

```

Now we have a data set with one day's worth of activity intensity values across each of the four activity intensity variables for each day (observation).

## Merging 
```{r}

```

## Aggregating & Summarizing 
```{r}

```

## Deriving 
```{r}

```
