---
title: "r4bs: Reading Data"
output: github_document
---
First things first - let's make sure you have all the necessary files. 

1. Navigate to the [r4bs github repo](https://github.com/erramirez/r4bs)
2. Dow load or Clone the repo
3. Open the r4bs.Rproj file

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading in Data
We're going to start with reading in data. This short tutorial will focus on .csv files as that tends to be a very common and standard format. 

### Base R and read.csv
We want to first explore how to use base R to read in a csv. We'll start with reading in the first file from `ExampleData\FitbitDailyData`. 

```{r read_base}
# use read.csv to read in the p01_dailyActivity.csv file and name it p01_base
# NOTE: using the full directory path here
p01_base <- read.csv("~/GitHub/r4bs/ExampleData/FitbitDailyData/p01_dailyActivity.csv")
```

Great! We created a data frame. What if we want to understand the structure of the data? We can use the `str()` function from the included `utils` package. 
```{r data_structure}
str(p01_base)
```
As you can see `str()` prints the structure of our data frame, giving us information on:
* the number of observations (rows)
* the number of variables (columns)
* the format or column type
 
While that _looks good_ I don't think that we actually want to store our `ActivityDate` variable as a factor. This is a quirk of the `read.csv` function - it sometimes converts strings to factors! So, let's dive into the Tidyverse and see how that works.

### Tidyverse, readr, and read_csv
The tidyverse loads the `readr` package, which is a powerful package for reading in data. The nice thing about readr is that it creates a `tibble` which is a bit different than a `data.frame`. One of the nice things that read_csv will do is _NOT_ change strings to factors! Let's see it in action 

```{r load_tidy}
# load tidyverse because it's great
library(tidyverse)
```

```{r read_date}
# use read_csv to read in the p01_dailyActivity.csv file and name it p01_readr
p01_readr <- read_csv("~/GitHub/r4bs/ExampleData/FitbitDailyData//p01_dailyActivity.csv")
```

Well look at that! The `read_csv` package not only loads the data but it automatically prints the column (variable) format. You'll also notice that `ActivityDate` is not a factor, but a character. Awesome. 

## Formatting Data
So now that we've used `read_csv` let's do some more work to better set up our data when we read it in. You may have noticed that the `ActivityDate` variable was in fact a date. While `readr` can detect some date and time formats it doesn't automatically parse this particular format (m/d/yy), and read it in as a character vector. 

When you're reading in your data you may need to update, change, or remove columns. You can use a few simple arguments to format your data as it's read it. This is very useful if you already know the format and content of your data. 

You can use the `col_types` argument to format your columns (cols) as they're read in via `readr`. There are a few different `col` arguments that allow you to specify the format of the column. Let's focus on `ActivityDate` here. 

```{r read_format}
# we use col_types to specify column formats
# we can either specifiy each column or just "edit" a specific column
# when editing one column all other column format are guessed using the normal parser

# NOTE: we use %m/%d/%y to indicate what format to parse
# NOTE: lower case y is used for two digit year, Y for four digit year
p01_readrformat <- read_csv("~/GitHub/r4bs/ExampleData/FitbitDailyData/p01_dailyActivity.csv",
                            col_types = cols(
                              ActivityDate = col_date("%m/%d/%y") 
                            )
)

head(p01_readrformat)
```

You can also do other types of formatting and data editing within your `read_csv` call. For example, let's say that we don't care about any of the distance information in this file. We can specify that we want to skip reading those in. 

```{r read_format_skip}
# you can also use other arguments to help with reading in data in a way that works for you
# we use col_skip here to specificy that we don't want any distance related data read in

p01_readrformatskip <- read_csv("~/GitHub/r4bs/ExampleData/FitbitDailyData/p01_dailyActivity.csv",
                            col_types = cols(
                              ActivityDate = col_date("%m/%d/%y"),
                              TotalDistance = col_skip(),
                              TrackerDistance = col_skip(),
                              LoggedActivitiesDistance = col_skip(),
                              VeryActiveDistance = col_skip(),
                              ModeratelyActiveDistance= col_skip(),
                              LightActiveDistance = col_skip(),
                              SedentaryActiveDistance = col_skip()
                              )
                            )
head(p01_readrformatskip)
```

You may not always be working with .csv files so you'll be happy to know that `readr` is a powerful package with functions for reading in all sorts of data, supporting six additional file formats:

* `read_csv2()`: uses ; for the field separator and , for the decimal point (good for EU files).
* `read_tsv()`: tab separated files
* `read_delim()`: general delimited files
* `read_fwf()`: fixed width files
* `read_table()`: tabular files where columns are separated by white-space.
* `read_log()`: web log files

### Working with Factors
Since not all data is numerical (or integer), we should probably touch on how to deal with factors. Factors are important, especially in survey data, as they represent categorical data. Let's work with some demographic data to better understand factors.

We provide a demographic data file here: `ExampleDate/FitbitDemographics/demographics.csv`. Let's load it and take a peak at what we have.

```{r read_demographics}

demographics <- read_csv("~/GitHub/r4bs/ExampleData/FitbitDemographics/demographics.csv")

head(demographics)
```
So we have 30 observations in a 9-variable data set here. Of those variables six were parsed as character variables, but in reality these are categorical variables and we want them to be read in as factors. 

Before we go down that path, let's take a brief pause here to explain the different ways factors can be set in R. 

```{r factor_nolevel}
# factor, no level
colors <- factor(c("yellow", "indigo", "green", "blue", "orange", "red", "violet"))

levels(colors)
```
Here, R understands that there are four levels of the categorical vector we called colors. It also guessed at the levels, placing them in alphabetical order. When working with factors, you can also set the level to your specific needs. 

```{r factor_levels}
# factor, with levels
colors <- factor(c("yellow", "indigo", "green", "blue", "orange", "red", "violet"))

# pass an list, in the correct order
colors <- factor(colors, levels= c("red", "orange", "yellow", "green", "blue", "indigo", "violet"))
levels(colors)
```
So now say that these categories actually have a distinct order that is useful for passing to computation arguments such as `min()` or `max()`. We can easily add an argument to make that happen.

```{r factor_levelsordered}
# factor, with levels and ordered
colors <- factor(colors, levels= c("red", "orange", "yellow", "green", "blue", "indigo", "violet"), ordered = TRUE)

min(colors)
max(colors)
```
Great. Now that we have a feeling for how factors work in R. We can set those variables we wish to keep as categorical as we read in our data. 

When reading in a factor using the `col_factor` argument we have a few different options. Let's check them using the built R help tools.

```{r col_factor_help}
?col_factor
```


```{r read_factor}
# reading in categorical variables as factor
demographics_factor <- read_csv("~/GitHub/r4bs/ExampleData/FitbitDemographics/demographics.csv", 
                                col_types = cols(
                                  Gender = col_factor(levels = c("Female", "Male")),
                                  Ethnicity = col_factor(NULL),
                                  Education = col_factor(levels = 
                                                           c("Some High School",
                                                             "Some College or Vocational Training",
                                                             "Completed College or University",
                                                             "Completed Graduate or Professional Degree")),
                                  `Martial Status` = col_factor(levels = NULL), 
                                  `Work Status` = col_factor(levels = NULL),
                                  Income = col_factor(levels = c("$10,000-$19,000",
                                                                 "$20,000-$29,000",
                                                                 "$30,000-$39,000",
                                                                 "$40,000-$49,000",
                                                                 "$50,000-$59,000",
                                                                 "$60,000-$69,000",
                                                                 "$70,000-$79,000",
                                                                 "$80,000-$89,000",
                                                                 "$90,000-$99,000",
                                                                 "> $100,000"), 
                                                      ordered = TRUE
                                                      )
                                )
)

# check to see if factors are correctly applied
head(demographics_factor)

# check to see if ordinal factor for Income is correctly applied
min(demographics_factor$Income)

# let's not take into account that NA (missing) data point
min(demographics_factor$Income, na.rm = TRUE)

# use summary to check for number of participants per category for Education
summary(demographics_factor$Education)

```


## Working with Excel Files
If you're working with Excel files (first, sorry - Excel can be the worst), then you can use another tidyverse package, `readxl` to work with excel files. Let's go through a quick example:

```{r read_excel}
# readxl is not loaded by default, but it is installed as part of the tidyverse
# you will have to explicitly load it first
library(readxl)

# read in p01_dailyActivity
# use the read_xlsx function as our data is a .xlsx file
p01_excel <- read_xlsx("~/GitHub/r4bs/ExampleData/FitbitDailyData/p01_dailyActivity.xlsx")

head(p01_excel)
```

## Learning More 

You can learn more about using `readr` here: http://readr.tidyverse.org/

You can learn more about using `readxl` here: http://readxl.tidyverse.org/
