---
title: "Exercise One: Answers"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Read in the first five participants (p01 - p05).
#### Drop/remove the following variables: `TrackerDistance`, `SedentaryDistance`, and `Calories`.

#### Set variables in the appropriate format(s).
### Read in the demographics file and set appropriate variables as factors.

#### Drop/remove the following variables: `Ethnicity` and `Education`.
###Read in the goals file.

### Tip: Reading Multiple Files
When you have multiple files that share the same structure you don't have to always read them in one by one. Let's go through a quick example of how to do that. 

We'll use a combination of creating a list of the file names. Since we're using files in a subdirectory we use the full path and also make sure to store the full file path name. We then pass the list of file names use `map_df` from the `purrr` package to read the files in and bind them together. 

```{r read_multiplefiles}
dailyactivity_all <- list.files(path = "~/GitHub/r4bs/ExampleData/FitbitDailyData/",
                    pattern = "*.csv", # only look for .csv files
                    full.names = TRUE) %>% # use the full path name
  map_df(~read_csv(., 
                   col_types = cols( 
                     ActivityDate = col_date("%m/%d/%y")
                     )
                   )
         )

```

