---
title: "r4bs: Part One"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Reading in Data
We're going to start with reading in data. This short tutorial will focus on .csv files as that tends to be a very common and standard format. 

### Base R and read.csv
We want to first explore how to use base R to read in a csv. We'll start with reading in the first file from `ExampleData\FitbitDailyData`. 

```{r}
# use read.csv to read in the p01_dailyActivity.csv file and name it p01_base
p01_base <- read.csv("https://raw.githubusercontent.com/erramirez/r4bs/master/ExampleData/FitbitDailyData/p01_dailyActivity.csv")
```

Great! We created a data frame. What if we want to understand the structure of the data? We can use the `str()` function from the included `utils` package. 
```{r}
str(p01_base)
```
As you can see `str()` prints the structure of our dataframe, giving us information on:
* the number of observations (rows)
* the number of variables (columns)
* the format or column type
 
While that _looks good_ I don't think that we actually want to store our `ActivityDate` variable as a factor. This is a quirk of the `read.csv` function - it sometimes converts strings to factors! So, let's dive into the Tidyverse and see how that works.

### Tidyverse, readr, and read_csv
The tidyverse loads the `readr` package, which is a powerful package for reading in data. The nice thing about readr is that it creates a `tibble` which is a bit different than a `data.frame`. One of the nice things that read_csv will do is _NOT_ change strings to factors! Let's see it in action 

```{r loadtidy}
# load tidyverse because it's great
library(tidyverse)

# use read_csv to read in the p01_dailyActivity.csv file and name it p01_readr
p01_readr <- read_csv("https://raw.githubusercontent.com/erramirez/r4bs/master/ExampleData/FitbitDailyData/p01_dailyActivity.csv")

```
Well look at that! The `read_csv` package not only loads the data but it automatically prints the the column (variable) format. You'll also notice that `ActivityDate` is not a factor, but a character. Awesome. 



